The Conjunctive Disjunctive Graph Node Kernel for Disease Gene Prioritization

### 2017-10-11 Response_3.tex

\documentclass[11pt]{article}
\usepackage{color}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage[utf8x]{inputenc}
\title{Revision notes for paper\\``The Conjunctive Disjunctive Graph Node Kernel''}
\author{Dinh Tran Van, Alessandro Sperduti and Fabrizio Costa }
\begin{document}
\maketitle

We thank the reviewer for the reviews and helpful comments on our manuscript ``The Conjunctive Disjunctive Graph Node Kernel''. 
In the following, we provide a description of how each comment (in \emph{italic}) has been addressed in this version of the paper.

\section{Reviewer 1}
\subsection*{1}
\begin{itemize}
\item \textit{lines 10-15: if the first claim (bias toward high degree nodes for the heat kernel) is understandable, the others are not and I don't get how the diffusion kernel (for instance) solves this issue, since it is very similar to the previous one. An evaluation of how the kernels address this property should be included in the experimental section or the authors should cite a paper already studying this property on different kernels;}

\underline{Answer}:

\item \textit{lines 20-25: what does "process in an additive and independent fashion" mean in this context. Similarly as before, please justify your claim about the poor discriminative power of these kernels;
}

\underline{Answer}:

\item \textit{lines 146-150: The first claim is misleading. The problem is not that the degree of some nodes is high it is that the degree distribution is highly skewed (and you must include a reference stating this fact for biological networks, as well as describe the own networks on which you are working for this property). Also, the rest of the argument is very strange: how do you get that the probability to find matches decreases exponentially. If there is a mathematical argument behind it, please, detail it;}

\underline{Answer}:

\end{itemize}

\subsection*{2} \textit{The paper is oriented exclusively toward an given application in biology. However, it is not visible from the title of the article and methodological choices are seldom justified in this context. Gene labelling should be evaluated: for instance, it would have been helpful to have a descriptive evaluation of the results of those labels in the experiment section. For the topological labelling, what is the biological meaning of nodes with similar degree being more similar? (give a reference) I don't understand if the topological labels are considered as a discrete or numerical label? An example of the way it partitions the network would be useful. For the functional labelling, evaluate your clusters in a way. Do they somehow relate to the graph structure or not? Do they correspond to clusters enriched in some functions? How robust is this clustering to very incomplete information as found in GO databases? In a similar fashion, can you provide an evaluation of your
decomposition on the networks you are working with?}

\underline{Answer}:

\subsection*{3} \textit{The paper is extremely badly organized, with some notions (like conjunctive and disjunctive edges) used before they are defined. I suggest the following re-organization of the paper:}

\begin{enumerate}
\item \textit{Introduction (that must include a discussion about the problem at stake, gene prioritization)}

\underline{Answer}:

\item \textit{A new kernel (or whatever title you want more informative than "Material and methods" that is a title better suited for bioinformatics journals) with A. Notations B. Node graph kernels + their problems C. Network decomposition D. Your kernel (with the case of discrete and numerical labels separated)}

\underline{Answer}:

\item \textit{Examples of node labelling in your context (to my opinion, Section 2.2.2 does not need the labels to be known. Just explain the kernel in Section 2 and then, in a separate section afterwards, explain your proposals for labelling with a better illustration of their usefulness in the biological framework)}

\underline{Answer}:

\item \textit{Experiments (since the discussion section is very short, I don't think that it is necessary or even relevant to have it separated from the rest of the paper)}
\end{enumerate}

\underline{Answer}:

\subsection*{4} \textit{The experimental part is very light. I would have expected to see an smart evaluation on a simulated dataset illustrating in which theoretical cases the approach is relevant. I think that an evaluation of the sensitivity of the method to some parameters is missing. For instance, why use a gene co-expression network with a correlation threshold of 0.5? (To my own experience, this is a very low threshold that should give a very dense network and is not the standard choices in this area). Also, why have twice less negative examples than positive ones (this is very far from being realistic since positive genes are usually very rare in this kind of real-life applications)?}

\underline{Answer}:

\section{Reviewer 2}
\begin{itemize}
\item \textit{I have found the paper to be hard to follow. This is mostly due to the large number of "sub-methods" that are involved in the proposed apparach, and to the too simplified description of the "global picture" of what the authors are proposing. It would be better to have an initial section or a table or a figure to summarize and easily introduce what the authors are going to propose/describe in the rest of the paper.}

\underline{Answer}:

\item \textit{Related to the previous comment, the notion of conjunctive and disjuctive edges seems central, however they are shortly mentioned in Section 2.2.2 and then shorlty re-defined (?) in Section 2.4. In the first case, the description of these edges must be expanced and clarified, togheter with some more details about Figure 1 (caption and text). Maybe it can be anticipated in the "Definitions" paragraph (page 3), or in the "initial decription" that I introduced in the previous comment.}

\underline{Answer}:

\item \textit{Overall, there is very large number of parameters involved in the proposed approach. The authors describe their model validation procedure in Page 9/10, however the space of possible combinations of parameters is huge + the choices over other thresholds described in the paper. This seems to be an important drawback of the method, and some comments on this point should be added.}

\underline{Answer}: The large number of parameters is a drawback of our propoased graph node kernel. However, some of them are suggested to have smale range of values to be effective.

\item \textit{Page 4, Functional label and real valued vector information encoding: The K-means algorithm is exploited, using the classical Euclidean distance, and the same distance is used in the vec.info.encoding. The authors should motivate this choice.}

\underline{Answer}: We have added a more detailed explanation in the new version of the paper.

\item \textit{Page 6, Section 2.3. A few words about the hash code computation can be helpful. }

\underline{Answer}: We have added a more detailed explanation in the new version of the paper.
\end{itemize}
\end{document}

### 2017-10-22 neurocomputing.tex
Add files via upload

\begin{abstract}
Biological relations encoded data are naturally represented in the form of graphs. Therefore, there is a high number of graph-based biological learning systems have been proposed and shown promissing results. One of the key points which determines the performance of graph-based learning systems is the denifition of node similarity measure. Node similarities are normally measured by graph node kernels. However, most available graph node kernels do not show high discriminative capacity since they share two common limitations. First they are based on the diffusion phenomenon which does not effectively exploit the nodes' configuration. Second they are not able to take into account the available information associated to nodes of graph.

In this paper, we propose an efficient graph node kernel that not only effectively exploits nodes' context, but also can integrate information which are available on the graph nodes. Empirical evaluation show that our proposed graph node kernel show state of the art concerning graph node kernels.

@@

\linenumbers
\section{Introduction}
Node similarity measure definition is the key that determines the performance of graph-based biological learning systems. The state of the art graph node kernels used to measure node similarity, are based on the notion of information diffusion, including LEDK \cite{kondor2002diffusion}, MEDK \cite{chen2014disease}, MDK \cite{fouss2006experimental}, RLK \cite{chebotarev2006matrix}, etc. These graph node kernels often show relatively low discriminative capacity, especially in cases of working with sparse graphs or graphs whose high numbers of missing links, since they share following limitations. First information is processed in an additive and independent fashion which prevents them from accurately modeling the configuration of each node's context. Second, they do not take into account information associated to nodes of graphs when they are available. These additional information normally provide a complement to graph topology. Therefore, they are potential to use to improve the expressiveness of graph node kernels.

@@

Predictive systems for gene-disease associations are often based on a notion of similarity between genes. A common strategy is to encode relations between genes as a network and to use graph based techniques to make useful inferences. In the last decades, a number of graph kernel methods have been proposed that directly exploit transitive properties in biological networks. The prototypical method is the Diffusion kernel (DK) \cite{kondor2002diffusion} inspired by the heat diffusion phenomenon. The key idea is to allow a given amount of {\em heat} placed on nodes to {\em diffuse} through the edges. The similarity between two nodes $v_{i}, v_{j}$ is then defined as the amount of heat starting from $v_{i}$ and reaching $v_{j}$ within a given time interval. In DK the heat flow is proportional to the number of paths connecting two nodes, which introduces an undesired bias that penalize peripheral nodes w.r.t. central ones. This problem is tackled by a modified version of DK called Markov exponential diffusion kernel (MED) \cite{chen2014disease} where a Markov matrix replaces the Laplacian matrix.  Another kernel called Markov diffusion kernel (MD) \cite{fouss2006experimental}, exploits instead the notion of {\em diffusion distances}, a measure of similarity between patterns of heat diffusion. The Regularized Laplacian kernel (RL) \cite{chebotarev2006matrix} represents instead a normalized version of the random walk with restart model and defines the node similarity as the number of paths connecting two nodes with different lengths. All these approaches can be applied to dense networks with high degree nodes. A drawback of these approaches is however their relatively low discriminative capacity. This is in part due to the fact that information is processed in an additive and independent fashion which prevents them from accurately modeling the configuration of each gene's context. To address this issue here we propose to employ a {\em decompositional} graph kernel (DGK) \cite{haussler1999convolution} technique in which the similarity function between graphs can be formed by decomposing each graph into subgraphs and by devising a valid local kernel between the subgraphs. To exploit its higher discriminative capacity we first decompose the network in a collection of connected sparse graphs and then we develop a suitable kernel, that we call Conjunctive Disjunctive Node Kernel (CDNK).

\cite{mordelet2011prodige}, \cite{chen2014disease}, \cite{valentini2014extensive}
\section{Background}
In this section, we first introduce definitions and notations that are used to define our proposed method. We then present graph node kernels and problems of state of the art graph node kernel.
\subsection{Definitions and Notations}
A graph is a structure, notated as $G=(\mathbb{V},\mathbb{E})$ where $\mathbb{V}$ is the set of nodes and $\mathbb{E}$ is the set of links (edges).

@@

\begin{definition}{}
\textit{Transition matrix of a graph $G$, notated as $P$, is a matrix whose each element $P_{ij} = A_{ij}/\sum_{i}^{}A_{ij}$ showing the probability of stepping on node $j$ from node $i$.}
\end{definition}

\subsection{Kernels defined on graphs}
Kernel methods, whose kernels, have emerged as one of the most powerful framework in machine learning which has been successfully applied in various fields. Kernels can be considered as similarity functions that allow to measure the similarity between any couple of objects in feature space and with any type of data representation.

@@

When the input of kernels is a set of graphs, we have graph kernels and when it is a set of graph nodes, we refer kernels as graph node kernels.
The problem of developing kernels defined on structured data in general and graphs in particular is made possible thanks to release of decomposition kernels proposed in \cite{haussler1999convolution}. Examples are kernels proposed in \cite{shervashidze2011weisfeiler, gartner2003survey, borgwardt2005shortest, costa2010fast}. Following we describe NSPDK \cite{costa2010fast}, a convolutional graph kernel which shows state of performance and it is later on used to define our proposed graph node kernel.

\subsubsection*{The Neighborhood Subgraph Pairwise Distance Kernel}
\label{sec:ndpdk}
The NSPDK \cite{costa2010fast} is an instance of convolution kernel \cite{haussler1999convolution} where given a graph $G \in \mathcal{G}$ and two rooted graphs $A_u, B_v$, the relation $R_{r,d}(A_u, B_v, G)$ is true {\em iff} $A_u \cong \mathcal{N}_r^u$ is (up to isomorphism $\cong$) a neighborhood subgraph of radius $r$ of $G$ and so is $B_v \cong  \mathcal{N}_r^v$, with roots at distance $\mathcal{D}(u,v)= d$. Figure \ref{sec:ndpdk} illustrates a pairwise neighborhood subgraph. We denote $R^{-1}$ as the inverse relation that returns all pairs of neighborhoods of radius $r$ at distance $d$ in $G$, $R^{-1}_{r,d}(G) = \lbrace A_u, B_v | R_{r,d}(A_u,B_v,G)=true\rbrace$. The kernel $\kappa_{r,d}$ over $\mathcal{G} \times \mathcal{G}$, counts the number of such fragments in common in two input graphs: 
\begin{center}
$\kappa_{r,d}(G,G^{'}) = 
\!\!\!\!\!\!\!\!\!\!\!\! 
\sum\limits_{\substack{A_u, B_v \ \in \ R_{r,d}^{-1}(G) \\ 
{A'}_{u'}, {B'}_{v'} \ \in \ R_{r,d}^{-1}(G')
}} \!\!\!\!\!\!\!\!\!\!\!\!  { { \textbf{1}_{A_{u} \cong A'_{u'}}} \cdot {
\textbf{1}_{B_{v} \cong B'_{v'}}} }$, 
\end{center} 
\noindent where $\textbf{1}_{A \cong B}$ is the \textit{exact matching function} that returns 1 if $A$ is
isomorphic to $B$ and 0 otherwise.  Finally, the NSPDK is defined as $K(G,G') = \sum\limits_{r}{\sum\limits_{d}{\kappa_{r,d}(G,G')}}$, where for efficiency reasons, the values of $r$ and $d$ are upper bounded to a given maximal $r^*$ and $d^*$, respectively.
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{nspdk}
\caption{Pairwise neighborhood subgraphs for the ``red'' node with $r=1$ and $d=3$}
\label{fig:nspdk}
\end{figure}

The state of the art kernels are based on diffusion phenomenon. In other words, they take paths connecting two nodes into account when measuring their similarity. In the following, we shortly introduce some of the most used graph node kernels.

\subsubsection{Laplacian exponential diffusion kernel}
One of the most well-known graph node kernels is the Laplacian exponential diffusion kernel \textbf{LEDK}, as it is widely used for exploiting discrete structures in general and graphs in particular. On the basis of the heat diffusion dynamics, Kondor and Lafferty proposed \textbf{LEDK} in \cite{kondor2002diffusion}: imagine to initialize each vertex with a given amount of heat and let it flow through the edges until an arbitrary instant of time. The similarity between any vertex couple $v_{i}$, $v_{j}$ is the amount of heat starting from $v_{i}$ and reaching $v_{j}$ within the given time. Therefore, \textbf{LEDK} can capture the long range relationship between vertices of a graph to define the global similarities. Below is the formula to compute \textbf{LEDK} values:
\begin{equation} 
\label{LEDK-formula}
K = e^{-\beta \textbf{L}} = \textbf{I} - \beta \textbf{L} + \frac{\beta \textbf{L}^{2}}{2!} - ...
\end{equation}
where $\beta$ is the diffusion parameter and is used to control the rate of diffusion and $\textbf{I}$ is the identity matrix. Choosing a consistent value for $\beta$ is very important: on the one side, if $\beta$ is too small, the local information cannot be diffused effectively and, on the other side, if it is too large, the local information will be lost. \textbf{LEDK} is positive semi-definite as proved in \cite{kondor2002diffusion}.
\subsubsection{Exponential diffusion kernel}
In \textbf{LEDK}, the similarity values between high degree vertices are generally higher compared to those between low degree ones. Intuitively, the more paths connect two vertices, the more heat can flow between them. This could be problematic since peripheral nodes have unbalanced similarities with respect to central nodes. In order to make the strength of individual vertices comparable, a modified version of \textbf{LEDK} is introduced by Chen et al in \cite{chen2014disease}, called Markov exponential diffusion kernel \textbf{MEDK} and given by the following formula:
\begin{equation} \label{MEDK-formula}
\textbf{K} = e^{-\beta \textbf{M}}
\end{equation}
The difference with respect to the Laplacian diffusion kernel is the replacement of \textbf{L} by the matrix $\textbf{M}=(\textbf{D}-\textbf{A}-n\textbf{I})/n$ where $n$ is the total number of vertices in graph. The role of $\beta$ is the same as for \textbf{LEDK}.
\subsubsection{Markov diffusion kernel}
The original Markov diffusion kernel \textbf{MDK} was introduced by Fouss et al. \cite{fouss2006experimental} and exploits the idea of diffusion distance, which is a measure of how similar the pattern of heat diffusion is among a pair of initialized nodes. In other words, it expresses how much nodes "influence" each other in a similar fashion. If their diffusion ways are alike, the similarity will be high and, vice versa, it will be low if they diffuse differently. This kernel is computed starting from the transition matrix \textbf{P} and by defining $\textbf{Z}(t) = \frac{1}{t}\sum_{\tau=1}^{t}\textbf{P}^{\tau}$, as follows:
\begin{equation} 
\label{MDK-formula}
\textbf{K} = \textbf{Z}(t)\textbf{Z}^{\top}(t)
\end{equation}

\subsubsection{Regularized Laplacian kernel}
Another popular graph node kernel function used in graph mining is the regularized Laplacian kernel \textbf{RLK}. This kernel function is introduced by Chebotarev and Shamis in \cite{chebotarev2006matrix} and represents a normalized version of the random walk with restart model. It is defined as follows:
\begin{equation} 
\label{RLK-formula}
\textbf{K} = \sum_{n=0}^{\infty}\beta^{n}(-\textbf{L})^n = (\textbf{I} + \beta \textbf{L})^{-1}
\end{equation}
where the parameter $\beta$ is again the diffusion parameter. \textbf{RLK} counts the paths connecting two nodes on the graph induced by taking \textbf{-L} as the adjacency matrix, regardless of the path length. Thus, a non-zero value is assigned to any couple of nodes as long as they are connected by any indirect path. \textbf{RLK} remains a relatedness measure even when diffusion factor is large, by virtue of the negative weights assigned to self-loops.

\section{Method}
We start from the type of similarity notion computed by a neighborhood based decomposition kernel between graph instances \cite{costa2010fast}, NSPDK, and adapt it to form a convolutional graph node kernel which aims at expressing the similarity between nodes in a single graph. Our intended graph node kernel takes an undirected, labeled graph as its input.

Given an input graph $G=(V,E)$, our method consists of three steps. In the first step, \textit{node labeling}, a labeling function is proposed to assign label for every node of the graph. This step is necessary since we desire to design a convolutional kernel which deals with labeled graphs. The second step includes a graph decompostion procedure which allows to decompose the graph into a set of linked sparse connected components. In this procedure, we define two different kinds of link: \textit{conjunctive} and \textit{disjunctive} in which they are treated in distinct manners later on. In the final step, the similarity between any node couple $u$ and $v$ is computed by adopting NSPDK on two neighborhood subgraphs rooted as $u$ and $v$ which extracted from the resulted graph after decomposing. Following, we describe each step in detail.

\subsection{Node Labeling} 
\label{sec:lab}
We are interested in investigating biological networks in general and genetic networks in specific. Therefore, in this section, we propose two different approaches to label for genetic graphs. It is worth to notice that genetic networks typically represent genes as nodes labeled with a gene identifier. Here we take a different approach and use the node labels to encode abstract information about the genes. In this way downstream machine learning algorithms can generalize from similar examples and allow the identification of overlooked but related genes. We experiment with two types of information: \textit{i}) topological information and \textit{ii}) functional information.\\

\textit{Topological labels} This information is simply based on the connectivity degree of the gene. The idea is that genes that have the same number of connections are more similar than genes with a different connectivity. Here we propose a node labeling function $\mathcal{L}$ that assigns labels for nodes by considering their degrees, such that nodes with similar degrees are assigned with a same label. Therefore, our function assigns the degree for nodes $u$ having degree less than or equal to a user defined threshold $T$ ($T=5$ in our experimental evaluation). However degree values larger than $T$ are subsequently discretized into $k$ levels. We treat nodes differently between nodes with high degrees and low degrees since we envisage that the properties of low-degree nodes whose different node degrees are more dissimilar in comparison with the ones of high degrees. Formally, the labeling function is defined as:
\begin{center}
$\mathcal{L}(u) = \left\{
	\begin{array}{ll}
		deg(u),\  & \mbox{if } deg(u) \leq T \\
		T+i,\ & \mbox{if } deg(u) > T
	\end{array}
\right.$,
\end{center}
where $i = \lceil \frac{deg(u)-T}{bin}\rceil$, $bin = \frac{deg(G)-T}{p - T}$ and $p$ ($p > T$) is the maximum number of symbols used. The value of $p$ depends on the degree distribution and can be tuned as a hyperparameter of the approach.\\

It is often that nodes in the graph are associated to extra information. These information can be considered as the complement to the graph topological information. Therefore, utilizing both information source are supposed to improve the performance of graph node kernels. In the following, we propose two different approaches to process with this kind of information. In the first approach, it is converted and used as the functional discrete labels for nodes, while it is processed and utilized as the real valued vectors in the second one. In both cases, we use \textit{Gene Ontology} \cite{gene2004gene} as an example of extra information associated to the graph nodes.\\

\textit{Functional discrete labels}
We use a gene ontology to construct binary vectors representing a bag-of-words encoding for each gene (\textit{i.e. each element of a binary vector is equal to 1 if its corresponding GO-term is associated to the gene, and is equal to 0 otherwise.}). The resulting vectors are then clustered using the k-means algorithm  into a user defined number of classes $K$ (tuned as a hyperparameter of the approach), so that genes with similar description profiles receive the same class identifier as label.\\

>>>>

\begin{abstract}
Biological relations encoded in data sources are naturally represented in the form of graphs. As a consequence, a high number of graph-based biological learning systems have been proposed and shown promissing results. In graph-based learning systems, one of the key points which determines their performance is the denifition of graph node similarity measure. Node similarities are normally measured by graph node kernels. However, most available graph node kernels do not show high discriminative capacity since they share two common limitations. First they are based on the diffusion phenomenon which does not effectively exploit the nodes' configuration. Second they are not able to proceed the auxiliary information associated to graph nodes.

In this paper, we propose an efficient graph node kernel that not only effectively exploits nodes' context, but also process with additional information which available on the graph nodes. Empirical evaluation on disease gene prioritization show that our proposed graph node kernel shows the state of the art concerning graph node kernels.

@@

\linenumbers
\section{Introduction}
The release of advanced technologies is one of the main reasons for the
revolution in various scientific research fields. In Biological and Medical
domain, modern technologies are making it not only easier but also more
economical than ever to undertake experiments and creating applications.
As a consequence, a vast amount of biological data in terms of volume
and type is generated through scientific experiments, published literature,
high-throughput experiment technology, and computational analysis. This
huge quantity of data are saved as biological datasets and made discoverable
through web browsers, application programming interfaces, scalable search
technology and extensive cross-referencing between databases. Biological
databases normally contain information about gene function, structure, lo-
calization, clinical effects of mutations and similarities of biological sequences and structures.

In Biomedical, disease-gene association recovery is a major goal in molecular  that has received much attention from many researchers. Despite that fact that a big progress has been made in the last decades, a number of genes known to be related to a genetic disease is normally limited. In order to find out the complement set of the known disease gene set, one way is to search for whole genome or specific regions that often contain a large number of suspected genes (candidate genes). This is obvious not a good idea as it is expensive not only in term of time consuming but also from financial aspect. For this reason, a considerable number of gene prioritization methods have been proposed. Predictive systems for gene-disease associations are often based on a notion of similarity between genes. A common strategy is to encode relations between genes as a network and to use graph based techniques to make useful inferences \cite{mordelet2011prodige}, \cite{chen2014disease}, \cite{valentini2014extensive}.

One of the key points that determines the performance of graph-based biological learning systems in general and graph-based disease gene prioritization systems in particular is the definition of node similarity measure. The node similarity is often measured by graph node kernels \cite{kondor2002diffusion}, \cite{chen2014disease}, \cite{fouss2006experimental}, \cite{chebotarev2006matrix}. However, the state of the art graph node kernels used to measure node similarity, are based on the notion of information diffusion. These graph node kernels often show relatively low discriminative capacity, especially in cases of working with sparse graphs, graphs have high numbers of missing links, since they share following limitations. First information is processed in an additive and independent fashion which prevents them from accurately modeling the configuration of each node's context. Second, they do not take into account information associated to nodes of graphs when they are available. These additional information normally provide a complement to graph topology. Therefore, they are potential to use to improve the expressiveness of graph node kernels.

@@

\section{Background}
In this section, we first introduce definitions and notations that are used to define our proposed method. We then describe the state-of-the-art concerning graph node kernels.
\subsection{Definitions and Notations}
A graph is a structure $G=(\mathbb{V},\mathbb{E}, \mathcal{L}_1, \mathcal{L}_2)$ where $\mathbb{V}$, $\mathbb{E}, \mathcal{L}_1, \mathcal{L}_2$ are the vertex (node) set, link (edge) set, discrete labeling fuction and real vector labeling function, respectively. The functions $\mathcal{L}_1, \mathcal{L}_2$ are defined as:
\begin{itemize}
\item $\mathcal{L}_1: \mathbb{V} \longmapsto \mathbb{L}$, where $\mathbb{L}$ is a set of discrete labels. $\mathcal{L}_1$ assigns a single discrete label $\ell \in \mathbb{L}$ for each ndoe $v \in \mathbb{V}$, $\mathcal{L}_1(v) = \ell$. 
\item $\mathcal{L}_2: \mathbb{V} \longmapsto \mathbb{R}^n$. $\mathcal{L}_2$ assigns a single real vector label $(v_1,v_2,\ldots,v_n) \in \mathbb{R}^n$ for each ndoe $v \in \mathbb{V}$, $\mathcal{L}_2(v) = (v_1,v_2,\ldots,v_n)$.
\end{itemize}

@@

\begin{definition}{}
\textit{Transition matrix of a graph $G$, notated as $P$, is a matrix whose each element value, $P_{ij} = A_{ij}/\sum_{i}^{}A_{ij}$, is proportional to the probability of stepping on node $j$ from node $i$.}
\end{definition}

\subsection{Kernels on graphs}
Kernel methods have emerged as one of the most powerful framework in machine learning, which has been successfully applied in various domains, due to their modularity. In kernel methods, the definition of kernel functions is independent from the design of the learning algorithm. A kernel function can be  considered as the similarity measure between example couples in the feature space. Interestingly, It can be defined in feature space and with any type of data representation.

@@

Canonical machine learning methods take vectorial data, data that are represented by vectors of features, as their input. However, there are many fields where data are not naturally represented by vectors, but structured forms in which graph is one of the most popular representation. Therefore, the task of developing methods which are able to learn from structured data in general or graphs in particular is very important. In this thesis, we focus on graphs, a special type of structured data representation.

The problem of developing kernels defined on structured data in general and graphs in particular is made possible thanks to the release of decomposition kernels proposed in \cite{haussler1999convolution}. If the input of a kernel is a set of graphs, it is called a graph kernel and if it is a set of graph nodes, the kernel is referred as graph node kernel. Both graph kernels and graph node kernels are widely applied to build graph-based learning systems ranging from social networks, to recommendation systems, to biology. In the following, we shortly describe some of the kernels in each category.

%Examples are kernels proposed in \cite{shervashidze2011weisfeiler, gartner2003survey, borgwardt2005shortest, costa2010fast}. Following we describe NSPDK \cite{costa2010fast}, a convolutional graph kernel which shows state of performance and it is later on used to define our proposed graph node kernel.
\subsubsection{Graph kernels}
\label{graph-kernels}
The task of designing effient and expressive graph kernels play an important role in the development of graph-based predictive systems. Existing graph kernels are decompositional kernels and can be classified into two categories: sequence-based graph kernels and subgraph-based graph kernels. The sequence-based graph kernels decompose graphs into ``parts" in sequence-based forms, such as paths and walks. Typical examples of sequence-based graph kernels are product graph kernel, \cite{gartner2003survey}, shortest path kernels \cite{borgwardt2005shortest}. The subgraph-based graph kernels dissolve graphs into subgraphs. Examples include Weisfeiler-Lehman kernels \cite{shervashidze2009fast, shervashidze2011weisfeiler}, The Neighborhood Subgraph Pairwise Distance Kernel (NSPDK) \cite{costa2010fast}. Following we describe NSPDK since it is later adopted to develop our graph node kernel (presented in \ref{method}).

The NSPDK \cite{costa2010fast} is an instance of convolution kernel \cite{haussler1999convolution} where given a graph $G \in \mathcal{G}$ and two rooted graphs $A_u, B_v$, the relation $R_{r,d}(A_u, B_v, G)$ is true {\em iff} $A_u \cong \mathcal{N}_r^u$ is (up to isomorphism $\cong$) a neighborhood subgraph of radius $r$ of $G$ and so is $B_v \cong  \mathcal{N}_r^v$, with roots at distance $\mathcal{D}(u,v)= d$. Figure \ref{sec:ndpdk} illustrates a pairwise neighborhood subgraph. We denote $R^{-1}$ as the inverse relation that returns all pairs of neighborhoods of radius $r$ at distance $d$ in $G$, $R^{-1}_{r,d}(G) = \lbrace A_u, B_v | R_{r,d}(A_u,B_v,G)=true\rbrace$. The kernel $\kappa_{r,d}$ over $\mathcal{G} \times \mathcal{G}$, counts the number of such fragments in common in two input graphs: 
\begin{center}
$\kappa_{r,d}(G,G^{'}) = 
\!\!\!\!\!\!\!\!\!\!\!\! 
\sum\limits_{\substack{A_u, B_v \ \in \ R_{r,d}^{-1}(G) \\ 
{A'}_{u'}, {B'}_{v'} \ \in \ R_{r,d}^{-1}(G')
}} \!\!\!\!\!\!\!\!\!\!\!\!  { { \textbf{1}_{A_{u} \cong A'_{u'}}} \cdot {
\textbf{1}_{B_{v} \cong B'_{v'}}} }$, 
\end{center} 
\noindent where $\textbf{1}_{A \cong B}$ is the \textit{exact matching function} that returns 1 if $A$ is
isomorphic to $B$ and 0 otherwise.  Finally, the NSPDK is defined as $K(G,G') = \sum\limits_{r}{\sum\limits_{d}{\kappa_{r,d}(G,G')}}$, where for efficiency reasons, the values of $r$ and $d$ are upper bounded to a given maximal $r^*$ and $d^*$, respectively.
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{nspdk}
\caption{Pairwise neighborhood subgraphs for the ``red'' node with $r=1$ and $d=3$}
\label{fig:nspdk}
\end{figure}

\subsubsection{Graph node kernels}
The state of the art kernels are based on diffusion phenomenon. In other words, they take paths connecting two nodes into account when measuring their similarity. In the following, we shortly introduce some of the most used graph node kernels.

\begin{itemize}

\item Laplacian exponential diffusion kernel: One of the most well-known graph node kernels is the Laplacian exponential diffusion kernel LEDK, as it is widely used for exploiting discrete structures in general and graphs in particular. On the basis of the heat diffusion dynamics, Kondor and Lafferty proposed LEDK in \cite{kondor2002diffusion}: imagine to initialize each vertex with a given amount of heat and let it flow through the edges until an arbitrary instant of time. The similarity between any vertex couple $v_{i}$, $v_{j}$ is the amount of heat starting from $v_{i}$ and reaching $v_{j}$ within the given time. Therefore, LEDK can capture the long range relationship between vertices of a graph to define the global similarities. Below is the formula to compute LEDK values:
\begin{equation} 
\label{LEDK-formula}
K = e^{-\beta L} = I - \beta L + \frac{\beta L^{2}}{2!} - ...
\end{equation}
where $\beta$ is the diffusion parameter and is used to control the rate of diffusion and $I$ is the identity matrix. Choosing a consistent value for $\beta$ is very important: on the one side, if $\beta$ is too small, the local information cannot be diffused effectively and, on the other side, if it is too large, the local information will be lost. LEDK is positive semi-definite as proved in \cite{kondor2002diffusion}.

\item Exponential diffusion kernel: In LEDK, the similarity values between high degree vertices are generally higher compared to those between low degree ones. Intuitively, the more paths connect two vertices, the more heat can flow between them. This could be problematic since peripheral nodes have unbalanced similarities with respect to central nodes. In order to make the strength of individual vertices comparable, a modified version of LEDK is introduced by Chen et al in \cite{chen2014disease}, called Markov exponential diffusion kernel MEDK and given by the following formula:
\begin{equation} \label{MEDK-formula}
K = e^{-\beta M}
\end{equation}
The difference with respect to the Laplacian diffusion kernel is the replacement of $L$ by the matrix $M=(D-A-nI)/n$ where $n$ is the total number of vertices in graph. The role of $\beta$ is the same as for LEDK.

\item Markov diffusion kernel: The original Markov diffusion kernel MDK was introduced by Fouss et al. \cite{fouss2006experimental} and exploits the idea of diffusion distance, which is a measure of how similar the pattern of heat diffusion is among a pair of initialized nodes. In other words, it expresses how much nodes "influence" each other in a similar fashion. If their diffusion ways are alike, the similarity will be high and, vice versa, it will be low if they diffuse differently. This kernel is computed starting from the transition matrix $P$ and by defining $Z(t) = \frac{1}{t}\sum_{\tau=1}^{t}P^{\tau}$, as follows:
\begin{equation} 
\label{MDK-formula}
K = Z(t)Z^{\top}(t)
\end{equation}

\item Regularized Laplacian kernel: Another popular graph node kernel function used in graph mining is the regularized Laplacian kernel RLK. This kernel function is introduced by Chebotarev and Shamis in \cite{chebotarev2006matrix} and represents a normalized version of the random walk with restart model. It is defined as follows:
\begin{equation} 
\label{RLK-formula}
K = \sum_{n=0}^{\infty}\beta^{n}(-L)^n = (I + \beta L)^{-1}
\end{equation}
where the parameter $\beta$ is again the diffusion parameter. RLK counts the paths connecting two nodes on the graph induced by taking $-L$ as the adjacency matrix, regardless of the path length. Thus, a non-zero value is assigned to any couple of nodes as long as they are connected by any indirect path. RLK remains a relatedness measure even when diffusion factor is large, by virtue of the negative weights assigned to self-loops.

\end{itemize}
\section{Method}
\label{method}
We start from the type of similarity notion computed by a neighborhood based decomposition kernel between graph instances \cite{costa2010fast}, NSPDK, and adapt it to form a convolutional graph node kernel which aims at expressing the similarity between nodes in a single graph. Our intended graph node kernel takes an undirected, labeled graph as its input.

Given an input labeled graph $G=(V,E, \mathcal{L}_1, \mathcal{L}_2)$, our kernel consists of two phases. In the first phase, a network decomposition procedure is applied to transform the graph into a set of linked sparse connected components. In this procedure, we define two different kinds of link: \textit{conjunctive} and \textit{disjunctive} in which we treat them in distinct manners. In the second step, the similarity between any node couple ($u$, $v$) is computed by adopting NSPDK on two neighborhood subgraphs rooted as $u$ and $v$ which extracted from the resulted graph after decomposing. In the following, we describe each phase in detail.

\subsection{Network Decomposition} 
\label{network-decomposition}
In genetic networks, it is not uncommon to find nodes with high degrees. Unfortunately these cases cannot be effectively processed by a neighborhood based decomposition kernel (see \ref{graph-kernels}) since these are based on the notion of exact matches. Neighborhood subgraphs rooted at high degree nodes are relatively big due to a high number of neighbors and edges. It leads to a low likelihood of finding neighborhood subgraphs rooted at high degree nodes which are isomorphic. Thus, the probability of having identical neighborhoods decreases exponentially as the degree increases. This means that in a finite network it quickly becomes impossible to find any match and hence learn or generalize at all.

We propose a procedure to ``sparsify'' the network that is observed by the neighborhood kernel. In practice, we mostly keep the same the cardinality of the edge set. However we mark the edges with special attributes so that kernel is able to treat them differently when computing. The result is a procedure that decomposes the network in a linked collection of sparse sub-networks where each node has a reduced connectivity when considering the edges of a specific type. However the other edges are still available to connect the various sub-networks. We distinguish two types of edges: {\em conjunctive} and {\em disjunctive} edges. Nodes linked by conjunctive edges are going to be used jointly to define the notion of context and will be visible to the neighborhood graph kernel. Nodes linked by disjunctive edges are instead used to define features based only on the pairwise co-occurrence of the genes at the endpoints and are processed by our novel kernel.

\textit{Iterative k-core decomposition} \cite{alvarez2005k}: the node set is partitioned in two groups on the basis of the degree of each node w.r.t. a threshold degree $D$, the first part contains all nodes with degree smaller than or equal to $D$ and the second part the remaining ones. The node partition is used to induce the ``conjunctive'' vs ``disjunctive'' notion for the edge partition: edges that have both endpoints in the same part are marked as conjunctive, otherwise they are marked as disjunctive. We apply the k-core decomposition iteratively, where at each iteration we consider only the graph induced by the conjunctive edges. We stop iterating the decomposition after a user defined number of steps. Note that this decomposition does not alter the cardinality of the edge set, it is simply a procedure to mark each edge with the attribute conjunctive or disjunctive.
